---
title: "Earthquake Analysis of Türkiye (2020-2026)"
author: "Feyza Ünal"
format: html
output-file: index.html
---

Türkiye is located on major fault lines and is highly prone to seismic activity. This project analyzes earthquake data between 2020 and 2026 using data obtained from AFAD. The aim is to identify temporal trends and spatial patterns in earthquake occurrences.

# 1. Project Overview and Scope

In this project, I analyze earthquake data from Türkiye obtained from AFAD (Disaster and Emergency Management Presidency) for the years 2020-2026.

My aim is to show, using analytical methods, how seismic activity has changed, particularly after the 2023 Kahramanmaraş earthquakes, and in which regions it has clustered.

# 2. Data

The dataset was provided by AFAD and consists of 9 columns.

The key columns are: `Magnitude`, `Latitude`, `Longitude`, and `Date`.

## 2.1 Data Source

https://deprem.afad.gov.tr/event-catalog

## 2.2 General Information About Data

```{r data-info}
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(readxl)

# ham veri okunur (filtre yok)
deprem_data_raw <- read_excel("~/Desktop/R/data.xlsx")

cat(
  "This dataset contains earthquake records in Türkiye between 2020 and 2026, sourced from AFAD. ",
  "The raw dataset includes ",
  nrow(deprem_data_raw),
  " observations and ",
  ncol(deprem_data_raw),
  " variables."
)

```

## 2.3 Reason of Choice

Türkiye is a country deeply affected by seismic activities due to its geographical location on major fault lines. The 2023 Kahramanmaraş earthquakes highlighted the importance of data-driven disaster management. I chose this dataset to analyze earthquake patterns using analytical tools and visualizations.

## 2.4 Preprocessing

The raw data required several preparation steps before analysis:

- The dataset was imported from an Excel file using the `readxl` library.
- The `Date` column was converted into `POSIXct` format.
- The `Magnitude` column was converted into numeric format.
- Only earthquakes between 2020 and 2026 were included.
- Only earthquakes with magnitude greater than 2 were included.

```{r data-preparation}
#| echo: false
#| message: false
#| warning: false
#| results: hide
library(readxl)
library(dplyr)

deprem_data <- read_excel("~/Desktop/R/data.xlsx")

# Date parse
deprem_data$Date <- as.POSIXct(
  deprem_data$Date,
  format = "%d/%m/%Y %H:%M:%S"
)

# Year çıkar
deprem_data$Year <- as.numeric(format(deprem_data$Date, "%Y"))

# Magnitude numeric
deprem_data$Magnitude <- as.numeric(deprem_data$Magnitude)

deprem_data <- deprem_data %>%
  filter(
    Year >= 2020,
    Year <= 2026,
    !is.na(Magnitude),
    Magnitude > 2   # ❗ 3 yerine 2
  )
```

After preprocessing, the final dataset contains 25,668 observations.

## 2.5 Global Data

In order to compare seismic activity in Türkiye with global trends, earthquake data from the United States Geological Survey (USGS) was used.

The global dataset includes earthquake records between 2020 and 2026. Relevant variables such as date, magnitude, latitude, and longitude were extracted and standardized to match the Türkiye dataset.

All datasets were combined into a single dataset with an additional variable indicating the region (Türkiye or World).


```{r global-data}
#| echo: false
#| message: false
#| warning: false

library(readr)
library(dplyr)

# Dünya verilerini oku
eq2020 <- read_csv("~/Desktop/R/usgs_2020.csv")
eq2021 <- read_csv("~/Desktop/R/usgs_2021.csv")
eq2022 <- read_csv("~/Desktop/R/usgs_2022.csv")
eq2023 <- read_csv("~/Desktop/R/usgs_2023.csv")
eq2024 <- read_csv("~/Desktop/R/usgs_2024.csv")
eq2025 <- read_csv("~/Desktop/R/usgs_2025.csv")
eq2026 <- read_csv("~/Desktop/R/usgs_2026.csv")

# Birleştir
world_data <- bind_rows(eq2020, eq2021, eq2022, eq2023, eq2024, eq2025, eq2026)

# 🔥 FORMATLA + FİLTRE (EN DOĞRU HAL)
world_data <- world_data %>%
  mutate(
    Date = as.POSIXct(time, format = "%Y-%m-%dT%H:%M:%OSZ", tz = "UTC"),
    Magnitude = as.numeric(mag),
    Latitude = as.numeric(latitude),
    Longitude = as.numeric(longitude),
    Region = "World"
  ) %>%
  filter(
    !is.na(Magnitude),
    Magnitude > 2   # 🔥 Türkiye ile aynı olmalı
  )

# Türkiye verisini etiketle
deprem_data <- deprem_data %>%
  mutate(Region = "Türkiye")

# Birleştir
combined_data <- bind_rows(deprem_data, world_data)
```


# 3. Analysis

## 3.1 Exploratory Data Analysis

In this section, I perform an initial exploration of the earthquake data by visualizing the distribution of earthquake magnitudes.

```{r violin-box}
#| echo: false
#| warning: false

library(dplyr)
library(ggplot2)

combined_data <- combined_data %>%
  mutate(Year = as.integer(format(Date, "%Y")))

ggplot(combined_data, aes(x = Region, y = Magnitude, fill = Region)) +
  geom_violin(alpha = 0.4, trim = FALSE) +
  geom_boxplot(width = 0.15, outlier.alpha = 0.2, color = "black") +
  facet_wrap(~Year) +
  scale_fill_manual(values = c("Türkiye" = "#E69F00", "World" = "#0072B2")) +
  labs(
    title = "Earthquake Magnitude Distribution by Year and Region",
    x = "",
    y = "Magnitude"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

The combined violin and boxplot visualization provides a detailed view of the distribution of earthquake magnitudes across years and regions. While the median values remain relatively stable, the shape of the distributions reveals differences in variability and the presence of high-magnitude outliers in global seismic activity.

## 3.2 Trend Analysis

The frequency analysis provides a clearer understanding of temporal changes in seismic activity compared to magnitude-based metrics. While earthquake magnitudes remain relatively stable over time, the number of earthquake events shows more noticeable variation.

The results indicate that global earthquake activity remains relatively consistent across the observed period. In contrast, Türkiye exhibits a significant increase in earthquake frequency, particularly after 2023. This increase is likely associated with major seismic events and aftershock sequences in the region.

These findings suggest that earthquake frequency is a more sensitive indicator of temporal changes in seismic activity, whereas magnitude primarily reflects the physical characteristics of individual events. Therefore, frequency-based analysis offers a more meaningful perspective for understanding recent seismic trends.rison allows us to determine whether changes in Türkiye are part of a global trend or driven by regional seismic activity.


```{r frequency-trend}
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2)

freq_data <- combined_data %>%
  mutate(Year = as.integer(format(Date, "%Y"))) %>%
  count(Region, Year)

ggplot(freq_data, aes(x = Year, y = n, color = Region, group = Region)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = 2020:2026) +
  scale_color_manual(values = c("Türkiye" = "#E69F00", "World" = "#0072B2")) +
  labs(
    title = "Earthquake Frequency Trend (2020–2026)",
    subtitle = "Comparison between Türkiye and Global Seismic Activity",
    x = "Year",
    y = "Number of Earthquakes",
    color = "Region"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```
```

The comparison shows that global earthquake activity remains relatively stable over time, while Türkiye exhibits a noticeable increase after 2023. This indicates that recent seismic activity in Türkiye is influenced by regional factors rather than global patterns.

## 3.3 Model Fitting

For the advanced analysis part, I may explore clustering methods such as K-means based on latitude and longitude to identify geographic hotspots of seismic activity. I may also examine simple time-based aggregations to better understand changes in earthquake frequency over time.

## 3.4 Results

Initial findings suggest that seismic activity in Türkiye is concentrated along major fault lines and that earthquake frequency varies across years.

## 3.5 Spatial Analysis

In this section, the spatial distribution of high-magnitude earthquakes is analyzed to identify geographic patterns and potential outliers that influence global seismic activity.

```{r map-outliers}
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)
library(maps)

outliers <- combined_data %>%
  filter(Magnitude >= 6)

ggplot(outliers, aes(x = Longitude, y = Latitude)) +
  borders("world", colour = "gray70", fill = "gray95") +
  geom_point(
    aes(size = Magnitude, color = Region),
    alpha = 0.7
  ) +
  scale_color_manual(values = c("Türkiye" = "#D55E00", "World" = "#0072B2")) +
  labs(
    title = "Global Distribution of High-Magnitude Earthquakes (M ≥ 6)",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

The spatial distribution of high-magnitude earthquakes shows that global seismic activity is concentrated along major tectonic boundaries, particularly in regions such as the Pacific Ring of Fire. These regions experience frequent high-magnitude earthquakes, which act as outliers and significantly influence the overall distribution of global earthquake magnitudes.

In contrast, Türkiye's seismic activity is more localized and primarily concentrated along major fault lines. Earthquakes in Türkiye are generally of moderate magnitude, which explains the differences observed between regional and global magnitude distributions.

# 4. Results and Key Takeaways

The frequency of earthquakes in Türkiye shows a noticeable increase after 2023.

Seismic activities are not uniformly distributed but are heavily clustered along the East and North Anatolian fault lines.

These patterns can provide useful insights for regional risk assessment and disaster preparedness.

### Data Download

Click the link below to download the processed dataset for reproducibility:

[Download deprem_verisi.RData](deprem_verisi.RData)


