---
title: "Earthquake Analysis of Türkiye (2020-2026)"
author: "Feyza Ünal"
format: html
output-file: index.html
---

Türkiye is located on major fault lines and is highly prone to seismic activity. This project analyzes earthquake data between 2020 and 2026 using data obtained from AFAD. The aim is to identify temporal trends and spatial patterns in earthquake occurrences.

# 1. Project Overview and Scope

In this project, I analyze earthquake data from Türkiye obtained from AFAD (Disaster and Emergency Management Presidency) for the years 2020-2026.

My aim is to show, using analytical methods, how seismic activity has changed, particularly after the 2023 Kahramanmaraş earthquakes, and in which regions it has clustered.

# 2. Data

The dataset was provided by AFAD and consists of 9 columns.

The key columns are: `Magnitude`, `Latitude`, `Longitude`, and `Date`.

## 2.1 Data Source

https://deprem.afad.gov.tr/event-catalog

## 2.2 General Information About Data

```{r data-info}
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(readxl)

# ham veri okunur (filtre yok)
deprem_data_raw <- read_excel("~/Desktop/R/data.xlsx")

cat(
  "This dataset contains earthquake records in Türkiye between 2020 and 2026, sourced from AFAD. ",
  "The raw dataset includes ",
  nrow(deprem_data_raw),
  " observations and ",
  ncol(deprem_data_raw),
  " variables."
)

```

## 2.3 Reason of Choice

Türkiye is a country deeply affected by seismic activities due to its geographical location on major fault lines. The 2023 Kahramanmaraş earthquakes highlighted the importance of data-driven disaster management. I chose this dataset to analyze earthquake patterns using analytical tools and visualizations.


## 2.4 Preprocessing

The raw data required several preparation steps before analysis:

- The dataset was imported from an Excel file using the `readxl` library.
- The `Date` column was converted into `POSIXct` format.
- The `Magnitude` column was converted into numeric format.
- Only earthquakes between 2020 and 2026 were included.
- Only earthquakes with magnitude greater than 2 were included.

```{r data-preparation}
#| echo: false
#| message: false
#| warning: false
#| results: hide
library(readxl)
library(dplyr)

deprem_data <- read_excel("~/Desktop/R/data.xlsx")

# Date parse
deprem_data$Date <- as.POSIXct(
  deprem_data$Date,
  format = "%d/%m/%Y %H:%M:%S"
)

# Year çıkar
deprem_data$Year <- as.numeric(format(deprem_data$Date, "%Y"))

# Magnitude numeric
deprem_data$Magnitude <- as.numeric(deprem_data$Magnitude)

deprem_data <- deprem_data %>%
  filter(
    Year >= 2020,
    Year <= 2026,
    !is.na(Magnitude),
    Magnitude > 2   # ❗ 3 yerine 2
  )
```

After preprocessing, the final dataset contains 25,668 observations.


## 2.5 Global Data

```{r global-data}
#| echo: false
#| message: false
#| warning: false

library(readr)
library(dplyr)

# Dünya verilerini oku
eq2020 <- read_csv("~/Desktop/R/usgs_2020.csv")
eq2021 <- read_csv("~/Desktop/R/usgs_2021.csv")
eq2022 <- read_csv("~/Desktop/R/usgs_2022.csv")
eq2023 <- read_csv("~/Desktop/R/usgs_2023.csv")
eq2024 <- read_csv("~/Desktop/R/usgs_2024.csv")
eq2025 <- read_csv("~/Desktop/R/usgs_2025.csv")
eq2026 <- read_csv("~/Desktop/R/usgs_2026.csv")

# Birleştir
world_data <- bind_rows(eq2020, eq2021, eq2022, eq2023, eq2024, eq2025, eq2026)

# 🔥 FORMATLA + FİLTRE (EN DOĞRU HAL)
world_data <- world_data %>%
  mutate(
    Date = as.POSIXct(time, format = "%Y-%m-%dT%H:%M:%OSZ", tz = "UTC"),
    Magnitude = as.numeric(mag),
    Latitude = as.numeric(latitude),
    Longitude = as.numeric(longitude),
    Region = "World"
  ) %>%
  filter(
    !is.na(Magnitude),
    Magnitude > 2   # 🔥 Türkiye ile aynı olmalı
  )

# Türkiye verisini etiketle
deprem_data <- deprem_data %>%
  mutate(Region = "Türkiye")

# Birleştir
combined_data <- bind_rows(deprem_data, world_data)
```

# 3. Analysis

## 3.1 Exploratory Data Analysis

In this section, I perform an initial exploration of the earthquake data by visualizing the distribution of earthquake magnitudes.

```{r}
#| echo: false

library(ggplot2)

ggplot(deprem_data, aes(x = Magnitude)) +
  geom_histogram(binwidth = 0.5, fill = "red", color = "white") +
  labs(
    title = "Magnitude Distribution of Earthquakes in Türkiye (2020-2026)",
    x = "Magnitude",
    y = "Count"
  )
```

## 3.2 Trend Analysis

This section examines how earthquake frequency has changed over time.

```{r trend-analysis}
#| echo: false

library(dplyr)
library(ggplot2)

trend_data <- combined_data %>%
  mutate(Year = format(Date, "%Y")) %>%
  count(Region, Year)

ggplot(trend_data, aes(x = Year, y = n, color = Region, group = Region)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Earthquake Trend: Türkiye vs World",
    x = "Year",
    y = "Number of Earthquakes"
  ) +
  theme_minimal()
```

## 3.3 Model Fitting

For the advanced analysis part, I may explore clustering methods such as K-means based on latitude and longitude to identify geographic hotspots of seismic activity. I may also examine simple time-based aggregations to better understand changes in earthquake frequency over time.

## 3.4 Results

Initial findings suggest that seismic activity in Türkiye is concentrated along major fault lines and that earthquake frequency varies across years.

# 4. Results and Key Takeaways

The frequency of earthquakes in Türkiye shows a noticeable increase after 2023.

Seismic activities are not uniformly distributed but are heavily clustered along the East and North Anatolian fault lines.

These patterns can provide useful insights for regional risk assessment and disaster preparedness.

### Data Download

Click the link below to download the processed dataset for reproducibility:

[Download deprem_verisi.RData](deprem_verisi.RData)


